### 2014-05-25 patchy-selection-paper.tex
edited application

%%%%%%%%%%
\section{Applications} 
Hoekstra, Nachman, and colleagues have made an extensive study of coat color in the rock pocket mouse, \emph{Chaetodipus intermedius}, building on classic work by \citet{dice1940ecologic,benson1933concealing}. 
These mice live on rock outcrops throughout the southwestern United States and northern Mexico, 
and generally have light pigmentation, presumably to match the color of the rock, avoiding visual-based predators. 
However, in some regions these mice live on darker substrates such as old lava flows, 
and in these patches have much darker pigmentation. 
\citet{nachman2003different} demonstrated that on one of these flows (Pinacate), an allele at the MC1R locus is responsible for much of the change to a dark pelage.
This dark allele differs from the light allele by four amino acid changes, 
and has a dominant or partially dominant effect depending on the measure of coat color assessed. 
These lava flows are separated from each other by a range of distances, 
with some being separated from any other by hundreds of kilometers of light colored substrate. 
The Pinacate allele is not present in a number of other populations of dark colored rock pocket mouse, 
suggesting these populations have adapted in parallel. 
However, there is some evidence \citep{Hoekstra:05} that, elsewhere in the range, 
multiple dark outcrops may share a dark phenotype whose genetic basis has been spread by migration
despite intervening light habitat. We use our framework to investigate
how the probability of parallel mutation depends on the distance
between lava flow patches.

As can be seen from the above, a key parameter is the characteristic length $\sigma/\sqrt{s_m}$. 
\citep{Hoekstra:04} studied the frequency of the dark MC1R allele, and coat color phenotypes, at sites on the Pinacate lava flow and at two nearby sites in light-colored rock.
On the lava flow the dark MC1R allele is at 86\% frequency. 
At the Tule light substrate site, which is $\sim$ 12km from the flow, the dark MC1R allele was at $\sim$ 3\% frequency. 
At a greater distance from the lava (XXXkm) the light colored substrate the dark MC1R was absent from a sample at the Christmas pass. 
While at the other light substrate site (O'Neill), which is 10km from the flow on opposite side from Tule, 
the dark MC1R allele has a frequency of $\sim$ 34\%. 
On the basis of these numbers we can gain a sense of a plausible range of characteristic lengths.
Assuming the dark allele is at 50\% frequency at the edge of the lava flow, 
we can fit the frequency found at each of the light substrate sites that are 
polymorphic for the dark allele under a simple allele frequency cline model \citep{XXX}.
Doing this for the Tule we obtain $\sigma/\sqrt{s_m} \approx 3$km and for the O'Neill site
$\sigma/\sqrt{s_m} \approx 30$km, giving us a range of cline widths. We also need estimate of $s_m$,
using  $\sigma=1km$ \gc{REF} our cline widths imply $s_m=3^{-1/2}$ and
$30^{-1/2}$ are reasonable values.

The mutational target size $\mu$ for the trait is unclear. 
While the Pinacate dark haplotype differs from the light haplotype at three amino-acid residues,
it is likely that not all of these changes are needed for a population to begin to  
adapt. Also there a number of genes, beyond MC1R, at which adaptative changes affecting 
pigmentation have been identified in closely related species and more
broadly across vertebrates \citep{REF}.To span a range of plausible
values we use a low mutation rate of $\mu= 10^{-8}$, where the adaptive
mtation has to occur at a single base pair; and a high mutation
rate $\mu= 10^{-5}$, equivalent to the mutation being able to arise at any of a thousand base pairs
e.g. any of the coding bases of a typical length gene.  Finally we set
$A=100km^2$ reflecting a patch with a width of $\sim 10km$. In Figure
\ref{fig:mice_prob_parallel} we show the dependance of the probability
of parallel mutation on the distance between lava flow patches using
these parameters. As can be seen parallel mutation should become
likely over a scale of tens to a few hundred kilometers between patches.

\begin{figure}[ht]
  \begin{center}
    \includegraphics{Lava_flow_mice_prob_parallel}
  \end{center}
  \caption{
The probability of parallel mutation as a function of the distance
between environmental patches for two different cline widths
and two different mutation rates (using eqn. \eqref{eqn:prob-parallel}). The parameters were chosen to match  
  \label{fig:mice_prob_parallel}
  }
\end{figure}

>>>>

%%%%%%%%%%
\section{Applications} 
Hoekstra, Nachman, and colleagues have made an extensive study of coat color in the rock pocket mouse, 
\emph{Chaetodipus intermedius}, building on classic work by \citet{benson1933concealing} and \citet{dice1940ecologic}.
These mice live on rocky outcrops throughout the southwestern United States and northern Mexico, 
and generally have a light pelage similar to the predominant rock color.
However, in some regions these mice live on darker substrates (e.g.\ old lava flows),
and in these patches have much darker pigmentation, 
presumably to avoid visual predators.
\citet{nachman2003different} demonstrated that on one of these flows (Pinacate), 
much of the change in coloration is due to an allele at the MC1R locus.
This dark allele differs from the light allele by four amino acid changes, 
and has a dominant or partially dominant effect depending on the measure of coat color assessed. 
These areas of dark rock are separated from each other by a range of distances, 
with some being separated from any other by hundreds of kilometers of light colored substrate. 
The Pinacate allele is not present in a number of other populations of dark colored rock pocket mouse, 
suggesting these populations have adapted in parallel \citep{xxx}.
However, there is some evidence \citep{Hoekstra:05} that, elsewhere in the range, 
multiple dark outcrops may share a dark phenotype whose genetic basis has been spread by migration
despite intervening light habitat. 
We apply our theory to this example,
in particular investigating how the probability of parallel mutation depends on 
the distance between lava flow patches.

A key parameter above was the dispersal distance divided by the square root of strength of selection against the focal allele between patches, $\sigma/\sqrt{s_m}$. 
\citep{hoekstra2004ecological} studied the frequency of the dark MC1R allele and coat color phenotypes, 
at sites across the (dark) Pinacate lava flow and at two nearby sites in light-colored rock.
On the lava flow the dark MC1R allele is at 86\% frequency,
while at a site with light substrate 12km to the west (Tule), the frequency is 3\%.
The dark allele was entirely absent from Christmas pass, a site with light substrate 7km north of Tule, and 3km further from the lava flow.
In the other direction, the dark MC1R allele was at 34\% at a site with light substrate 10km to the east of the flow (O'Neill).
These numbers give us a sense of a plausible range of characteristic lengths.
Assuming the dark allele is at 50\% frequency at the edge of the lava flow, 
we can fit formula \eqref{eqn:eqfreq} to these frequencies.
Doing this for Tule we obtain $\sigma/\sqrt{s_m} \approx 3$km, 
and for O'Neill, $\sigma/\sqrt{s_m} \approx 30$km, giving us a range of cline widths. 
We also need estimate of $s_m$. 
Using $\sigma=1km$ \gc{REF}, these cline widths imply $s_m=1/\sqrt{3}$ and
$1/\sqrt{30}$ are reasonable values.
\plr{Going back to previous conversations, we came upon \citet{mullen2008natural}, 
which has dispersal distance of 0.26km.  Shall we ask Nachman?}

The mutational target size $\mu$ for the trait is unclear. 
While the Pinacate dark haplotype differs from the light haplotype at three amino-acid residues,
it is likely that not all of these changes are needed for a population to begin to  adapt. 
Also there a number of genes besides MC1R at which adaptative changes affecting pigmentation 
have been identified in closely related species and more broadly across vertebrates \citep{REF}.
To span a range of plausible values, we use a low mutation rate of $\mu= 10^{-8}$, 
where the adaptive mtation has to occur at a single base pair; 
and a high mutation rate $\mu= 10^{-5}$, equivalent to the mutation being able to arise at any of a thousand base pairs,
e.g.\ any of the coding bases of a typical gene.  
Finally, we set $A=100\text{km}^2$.
Figure~\ref{fig:mice_prob_parallel} shows the dependance of the probability
of parallel mutation on the distance between lava flow patches using these parameters. 
As can be seen parallel mutation should become
likely over a scale of tens to a few hundred kilometers between patches.

\begin{figure}[ht]
  \begin{center}
    \includegraphics{Lava_flow_mice_prob_parallel}
  \end{center}
  \caption{
The probability of parallel mutation as a function of the distance
between environmental patches for two different cline widths
and two different mutation rates (using equation \eqref{eqn:prob-parallel}). 
The parameters were chosen to match  
  \label{fig:mice_prob_parallel}
  }
\end{figure}
